---
title: "Project"
author: "Eduardo Marenco"
subtitle: "Investigation into Chicago taxis"
echo: false
bibliography: references.bib
---

## Packages 

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(tidymodels)
library(scales)
```

First, I took a look to the data set to get a sense of the data.


View(taxi)

## Question 1

I used ggplot2 to visualize the relationship between variables tip and distance.

```{r}
#| label: load-packages
# Load ggplot2 for enhanced data visualization
library(ggplot2)
```

A box plot suggests that there may be a relationship between the distance traveled and tips granted. 

```{r}
#| label: distance-tip-boxplot
#| fig-cap: "Distribution of Trip Distance by Tipping Status"

# Create the Box Plot using ggplot2
ggplot(taxi, aes(x = tip, y = distance, fill = tip)) +
  geom_boxplot() +
  labs(
    title = "Trip Distance Distribution by Tipping Status",
    x = "Tip Status",
    y = "Distance (miles)"
  ) +
  theme_minimal()
```
A t test shows that the average distance for rides that received a tip was approximately 6.37 miles and that the the average distance for rides that did not receive a tip was approximately 4.57 miles. The mean distance for tipping rides is about 1.8 miles longer than non-tipping rides ($6.37 - 4.57$). 

```{r}
#| label: two-sample-t-test
# Perform the Two-Sample t-Test (Distance ~ Tip)
# The formula syntax is: <quantitative_variable> ~ <categorical_variable>
t_test_result <- t.test(distance ~ tip, data = taxi)

# Print the results of the t-test
print(t_test_result)
```
In conclusion, there is a statistically significant difference in the mean trip distance between rides that received a tip and those that did not. Customers are significantly more likely to leave a tip on longer trips compared to shorter ones. 

## Question 2 


## Your own question 


