---
title: "Project"
author: "Eduardo Marenco"
subtitle: "Investigation into Chicago taxis"
echo: false
bibliography: references.bib
execute: 
  warning: false
fig-width: 6
fig-asp: 0.618
---

## Packages 

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(tidymodels)
library(scales)
```

First, I took a look to the data set to get a sense of the data using View(taxi)

## Question 1

I used ggplot2 to visualize the relationship between variables tip and distance.

```{r}
#| label: load-packages
# Load ggplot2 for enhanced data visualization
library(ggplot2)
```

A box plot suggests that there may be a relationship between the distance traveled and tips granted. In fact, a density plot shows clearly that the distribution of tips concentrates below the 20 miles distance threshold as shown below where the peak for tipped trips (in red) is higher than for the trips with no tip status given.

```{r}
#| label: distance-tip-boxplot
#| fig-cap: "Distribution of Trip Distance by Tipping Status"

# Create the Box Plot using ggplot2
ggplot(taxi, aes(x = tip, y = distance, fill = tip)) +
  geom_boxplot() +
  labs(
    title = "Trip Distance Distribution by Tipping Status",
    x = "Tip Status",
    y = "Distance (miles)"
  ) +
  theme_minimal()
```
```{r}
#| label: distance-tip-density plot
#| fig-cap: "Distribution of Trip Distance by Tipping Status"
ggplot(
  taxi,
  aes(x = distance, color = tip, fill = tip)
) +
  geom_density(alpha = 0.5)
```



A t test shows that the average distance for rides that received a tip was approximately 6.37 miles and that the the average distance for rides that did not receive a tip was approximately 4.57 miles. The mean distance for tipping rides is about 1.8 miles longer than non-tipping rides ($6.37 - 4.57$). 

```{r}
#| label: two-sample-t-test
# Perform the Two-Sample t-Test (Distance ~ Tip)
# The formula syntax is: <quantitative_variable> ~ <categorical_variable>
t_test_result <- t.test(distance ~ tip, data = taxi)

# Print the results of the t-test
print(t_test_result)
```
In conclusion, there is a statistically significant difference in the mean trip distance between rides that received a tip and those that did not. Customers are significantly more likely to leave a tip on longer trips compared to shorter ones. 

## Question 2 


## Your own question 


